---
title: "MEASO Open biodiversity data gap analysis"
subtitle: "Results"
date: "`r format(Sys.time(), '%a %d %b %Y')`"
# author: "Anne-Sophie Bonnet-Lebrun"
output:
  pdf_document:
  always_allow_html: true
  # html_document:
  fig_caption: yes
---

```{r setup, include=FALSE}

library(knitr)
# knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

inline_hook <- function(x) {
  if (is.numeric(x)) {
    formatted <- format(x, big.mark = ",")
  } else {
    formatted <- x
  }
  return(formatted)
}

knit_hooks$set(inline = inline_hook)
```

```{r initiate_environment, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
# library(worms)
# library(worrms)
library(knitr)
library(data.table)

# https://github.com/AustralianAntarcticDivision/SOmap
# remotes::install_github("AustralianAntarcticDivision/SOmap")
library(SOmap)
library(ggnewscale)
library(ggpubr)
library(cowplot)
library(patchwork)
library(sf)

all_measo <- fread("../../data/edited/all_measo_land_worms_date_dupli.csv")

all_measo <- all_measo %>%
  dplyr::mutate(
    QC_basisOfRecord = tolower(basisOfRecord) %>% str_replace("_", ""),
    QC_basisOfRecord = ifelse(QC_basisOfRecord == "machineobservation", "machine", ifelse(basisOfRecord == "fossilspecimen", "fossil", "human"))
  )

```


# Data availability and overall quality

The search criteria matched with `r sum(all_measo$QC_database == "OBIS", na.rm = TRUE)` records on OBIS and `r sum(all_measo$QC_database == "GBIF", na.rm = TRUE)` on GBIF. Of these, `r sum(all_measo$QC_onLand, na.rm = TRUE)` records appeared to be situated on land (Fig. 1). In total, `r sum(all_measo$QC_hasScientificNameInWorms == FALSE, na.rm = TRUE)` records had no *scientificName* (`r round(sum(all_measo$QC_hasScientificNameInWorms == FALSE, na.rm=T)*100/nrow(all_measo), 2)`% of the data), and `r sum(all_measo$QC_hasScientificNameInWorms == "not in WoRMS", na.rm = TRUE)` records had a scientificName with no match in WoRMS (`r round(sum(all_measo$QC_hasScientificNameInWorms == "not in WoRMS", na.rm=T)*100/nrow(all_measo), 2)`% of the data). Note that `r round(sum(all_measo$QC_hasScientificNameInWorms == "not in WoRMS" & all_measo$organismQuantityType == "DNAsequencereads", na.rm = TRUE) / sum(all_measo$QC_hasScientificNameInWorms == "not in WoRMS", na.rm = TRUE) * 100, 2)` % if the records with no match in WoRMS come from DNA sequence reads.

`r round(sum(all_measo$QC_date %in% c("missing", "incomplete", "conflicting"), na.rm = TRUE)*100/nrow(all_measo),2)`% of records had incomplete date information (year missing for `r round(sum(is.na(all_measo$year2))*100/nrow(all_measo),2)`% of records, month for `r round(sum(is.na(all_measo$month2))*100/nrow(all_measo),2)`% and day for `r round(sum(is.na(all_measo$day2))*100/nrow(all_measo),2)`%). In addition, `r sum(all_measo$year < 1773 | all_measo$year2 < 1773, na.rm=TRUE)` records prior to 1773 (first recorded crossing of the Antarctic circle), and `r sum(all_measo$year > 2022 | all_measo$year2 > 2022, na.rm=TRUE)` records posterior to 2022, were flagged as erroneous (e.g. `r sum(all_measo$year == 0 | all_measo$year2 == 0, na.rm=TRUE)` records supposedly in year 0). `r sum(!all_measo$QC_notAbsent)` records (`r round(sum(!all_measo$QC_notAbsent)*100/nrow(all_measo),2)`% of all data) referred to species absences (all from GBIF; Fig. 1). 

`r sum(all_measo$QC_trueDuplicates, na.rm=T)` records (`r round(sum(all_measo$QC_trueDuplicates, na.rm=T)*100/nrow(all_measo), 2)`% of the data) were true duplicates of already existing records (`r sum(all_measo$QC_trueDuplicatesIntra[all_measo$QC_database == "OBIS"], na.rm=T)` within OBIS, `r sum(all_measo$QC_trueDuplicatesIntra[all_measo$QC_database == "GBIF"], na.rm=T)` within GBIF and `r sum(all_measo$QC_trueDuplicates, na.rm=T) - sum(all_measo$QC_trueDuplicatesIntra, na.rm=T)` between the two databases).
`r sum(all_measo$QC_potentialDuplicates_taxCheck, na.rm=T)` (`r round(sum(all_measo$QC_potentialDuplicates_taxCheck, na.rm=T)*100/nrow(all_measo), 2)`% of the total) records were potential duplicates of other records (`r sum(all_measo$QC_potentialDuplicatesIntra[all_measo$QC_database == "OBIS"], na.rm=T)` within OBIS, `r sum(all_measo$QC_potentialDuplicatesIntra[all_measo$QC_database == "GBIF"], na.rm=T)` within GBIF, and `r sum(all_measo$QC_potentialDuplicates_taxCheck, na.rm=T) - sum(all_measo$QC_potentialDuplicatesIntra, na.rm=T)` between the two databases, leaving `r sum(!all_measo$QC_potentialDuplicates_taxCheck)` unique records in the database. Machine observations accounted for `r round(sum(all_measo$QC_basisOfRecord == "machine",na.rm=T)*100/nrow(all_measo), 2)`% of all records (before quality controls), and human observations for `r round(sum(all_measo$QC_basisOfRecord == "human",na.rm=T)*100/nrow(all_measo), 2)`% (the rest being unclassified or fossil records).

Only `r round(sum(!is.na(all_measo$individualCount))*100/nrow(all_measo),2)`% of all (i.e., presence and absence) records contained information on the number of individuals observed (*individualCount* field), and `r round(sum(!is.na(all_measo$organismQuantity))*100/nrow(all_measo),2)`% of all records contained abundance information (i.e., explicit mention of number of individuals per unit of space and/or time in the *organismQuantity* and *organismQuantityType* fields). Note that absence records could be considered has carrying abundance information (i.e., null abundance) but the *individualCount* and *organismQuantity* fields were filled in for respectively only `r round(sum(!is.na(all_measo$individualCount) & !all_measo$QC_notAbsent)*100/sum(!all_measo$QC_notAbsent),2)`% and `r round(sum(!is.na(all_measo$organismQuantity) & !all_measo$QC_notAbsent)*100/sum(!all_measo$QC_notAbsent),2)`% of absence records. An occurrenceID was missing for `r sum(is.na(all_measo$occurrenceID))` records (`r round(sum(is.na(all_measo$occurrenceID))*100/nrow(all_measo),2)`%), probably pre-dating the implementation of *occurrenceID* as an obligatory field (`r round(sum(is.na(all_measo$occurrenceID) & all_measo$year2 < 2007, na.rm=TRUE)/sum(is.na(all_measo$occurrenceID), na.rm=TRUE)*100,2)`% of which collected prior to 2007). 

```{r, datasets}

datasets <- all_measo %>%
  group_by(datasetName) %>%
  dplyr::summarise(N = n())
```

Records originated from `r all_measo$datasetName %>% na.omit() %>% unique() %>% length()` individually identified datasets (based on unique value in the *datasetName* field) and `r all_measo$institutionCode %>% na.omit() %>% unique() %>% length()` institutions (*institutionCode* field) - but note that `r sum(is.na(all_measo$datasetName))` records had no data in the *datasetName* field and `r sum(is.na(all_measo$institutionCode))` records had no data in the *institutionCode* field. The contributions of individual datasets were highly variable: `r sum(datasets$N<10, na.rm=TRUE)` dataset names with less than 10 records, while `r sum(datasets$N>50000, na.rm=TRUE)` dataset names match with more than 50,000 records (including “The Retrospective Analysis of Antarctic Tracking (Standardised) Data from the Scientific Committee on Antarctic Research”, `r  filter(datasets, datasetName == "TheRetrospectiveAnalysisofAntarcticTracking(Standardised)DatafromtheScientificCommitteeonAntarcticResearch")$N` records; and “Southern Ocean Continuous Zooplankton Recorder (SO-CPR) Survey”, `r  filter(datasets, datasetName == "SouthernOceanContinuousZooplanktonRecorder(SO-CPR)Survey")$N` records).

```{r, remove datasets}

rm(datasets)
```


# Exploration of distributions and biases

All subsequent results will refer to the filtered dataset (no terrestrial records, presences only, unique records, complete and realistic dates, machine or human observations).

```{r, working dataset}

# change to update it in this document - or refer to a script that updates it
all_measo_wd <- data.table::fread("../../data/edited/all_measo_wd_land_worms_date_dupli_focalgps_cells.csv")

# peak in may in the last decade
may_2010_2019 <- 
  all_measo_wd %>% 
  filter(
    month == 5,
    QC_basisOfRecord == "human",
    between(year2, 2010, 2019)
    )
# one main campaign for eDNA in 2016
may_2010_2019_DNA <- may_2010_2019 %>% 
  filter(
    str_detect(QC_measo, "WP"),
    organismQuantityType == "DNAsequencereads",
    year2 == 2016
    )

```

## Temporal distribution
Most observations were collected during the 20th and 21st centuries (only `r round(sum(all_measo_wd$year2 < 1900, na.rm = TRUE)/nrow(all_measo_wd)*100,2)`% of records prior to 1900, `r round(sum(all_measo_wd$year2 < 1900 & (tolower(all_measo_wd$basisOfRecord) %>% str_replace("_", "")) == "preservedspecimen", na.rm = TRUE)/sum(all_measo_wd$year2 < 1900, na.rm = TRUE)*100,2)`% of which classified as preserved specimen), with an increase starting around the 1950ies and a decrease in the last decade – for both machine and human observations (Fig. 2A). There was also a consistent seasonal bias in the data, particularly for human observations, which showed higher sampling intensity during the austral summer months (Fig. 2B). Note that the peak in human observations in May in the last decade is dominated by what seems to be one campaign collecting and analysing environmental DNA (i.e., "DNA sequence reads" in the *OrganismQuantityType* column) in 2016 in the West Pacific MEASO sector (`r round(nrow(may_2010_2019_DNA)/nrow(may_2010_2019)*100, 2)` % of the 2000-2019 May peak).

## Geographic distribution
Human observations were highly clustered (Fig. 3A), with high record densities around Sub-Antarctic islands, and visible cruise tracks, e.g., between the Antarctic continent and New-Zealand or Australia. The number of human observation records per unit area was markedly highest – and increased the most in recent years – in the Sub-Antarctic zone of the Central Indian sector, the Antarctic zone of the East Pacific sector, and the sub-Antarctic and Northern zones of the East Indian sector (Fig. 4). High numbers of human observations were found around sub-Antarctic Islands, and along routes leading to research stations (Supplementary Fig. 1A). In winter, human observations were almost absent from large areas covered by sea ice (Supplementary Fig. 2), although areas in the Central Indian Ocean remained sampled in winter despite the presence of sea ice. In contrast, although most abundant around research stations (Antarctic peninsula, sub-Antarctic islands, some areas on the Antarctic coast; Supplementary Fig. 1B), machine observation data covered a larger part of the Southern Ocean and appeared not to be confounded by ship movement (Fig. 3B, Supplementary Figs. 1B and 3). However, in the case of tracking data from central-place foragers (a large part of the dataset), they will likely be biased by sampled colonies. 

## Taxonomic composition
Overall, the data included `r length(unique(all_measo$valid_name))` different taxonomic units, based on values from the *scientificName* field, after matching with the WoRMS taxonomic backbone. These units ranged from (sub-)species to kingdom-level identifications. 

```{r, phyla}

rm(may_2010_2019)
rm(may_2010_2019_DNA)

phyla <- all_measo_wd %>%
  filter(!is.na(phylum_worms)) %>%
  group_by(phylum_worms) %>%
  dplyr::summarise(N = n()) %>%
  arrange(desc(N))
```

After matching with the WoRMS taxonomic backbone, we identified `r length(unique(all_measo_wd$phylum_worms))` individual phyla in the data. The 10 best represented phyla for the whole working dataset (machine and human observations) were `r phyla$phylum_worms[1]` (`r phyla$N[1]` records), `r phyla$phylum_worms[2]` (`r phyla$N[2]`), `r phyla$phylum_worms[3]` (`r phyla$N[3]`), `r phyla$phylum_worms[4]` (`r phyla$N[4]`), `r phyla$phylum_worms[5]` (`r phyla$N[5]`), `r phyla$phylum_worms[6]` (`r phyla$N[6]`), `r phyla$phylum_worms[7]` (`r phyla$N[7]`), `r phyla$phylum_worms[8]` (`r phyla$N[8]`), `r phyla$phylum_worms[9]` (`r phyla$N[9]`), and `r phyla$phylum_worms[10]` (`r phyla$N[10]`). The evolution of the number of records through time varied across phyla, with several phyla increasing until 2001-2010 and then plateauing or decreasing in the last decade, while Proteobacteria, which were absent from the databases until the decade 1980-1989, showing a very marked increase continuing in the last decade, making them to the 3rd currently most-sampled phylum (Supplementary Fig. 4). 

The depth of taxonomic identification varied with observation types: while `r round(sum(!is.na(all_measo_wd$genus_worms) & all_measo_wd$QC_basisOfRecord == "machine", na.rm = TRUE)/sum(all_measo_wd$QC_basisOfRecord == "machine", na.rm = TRUE)*100, 2)`% of machine observations were identified at the genus level, and `r round(sum(!is.na(all_measo_wd$species_final) & all_measo_wd$QC_basisOfRecord == "machine", na.rm = TRUE)/sum(all_measo_wd$QC_basisOfRecord == "machine", na.rm = TRUE)*100, 2)`% at the species level, only `r round(sum(!is.na(all_measo_wd$genus_worms) & all_measo_wd$QC_basisOfRecord == "human", na.rm = TRUE)/sum(all_measo_wd$QC_basisOfRecord == "human", na.rm = TRUE)*100, 2)`% of human observations were identified at the genus level, and only `r round(sum(!is.na(all_measo_wd$species_final) & all_measo_wd$QC_basisOfRecord == "human", na.rm = TRUE)/sum(all_measo_wd$QC_basisOfRecord == "human", na.rm = TRUE)*100, 2)`% at the species level. The depth of taxonomic identification also varied among phyla, with large and morphologically easily identifiable phyla such as Chordata (but also Mollusca, up to genus level) identified more precisely than phyla such as Radiozoa, Chaetognatha, Foraminifera and Proteobacteria which require microscopy or DNA-based techniques to reach fine level identifications (Supplementary Fig. 5).

Overall, pelagic species were better sampled than benthic species (higher number of records and higher spatial coverage (Fig. 5). There were large disparities in the number of human observations covering each focal group (Fig. 5A), as well as in the spatial coverage of each group (Fig. 5B). Birds and mammals were by far the most sampled group, followed by Crustacea. Birds and mammals, and Crustacea were also the groups with the best spatial coverage. However, some groups had relatively low sampling but relatively good coverage (e.g., Mollusca, Gelatinous zooplankton, or Annelida). No group had a spatial coverage higher than 65%. Both the distribution of records among groups and spatial coverage also varied across MEASO areas (Supplementary Figs. 6 and 7).

In most sectors, no plateau was reached in the genus accumulation curves (Supplementary Figs. 8, 9 and 10), and even in the sector where a plateau was reached, there were very few cells that appeared to be sampled enough. In addition, the fit was poor for most sectors (including some cells above the prediction line).

# Suitability of the available data for a MEASO
Table 1 shows, for each relevant EBV or EOV, the data required for its calculation, and the fit for purpose of OBIS and GBIF data. In general, geo-referenced data are available for a range of taxa but would require taking the spatio-temporal biases identified above into account. For variables that require abundance or depth data, only small proportions of records contain this information. 


